---
title: "host_blood analysis"
output: html_notebook
---

This notebook assess mosquito blood meal sources 

Data description

An experiment was conducted in Tulizamoyo to to assess the proportion of two Anopheles mosquito Species: An. funestus and An.  arabiensis (species), blood feeding on vertebrates (host_blood). The experiment was done in a field setting, and mosquitoes were collected in different houses (hh_id), using different traps (trap_method), indoors and outdoors (position). PCR targeting 
cytocrome b was performed on the samples collected, and the number of amplified samples were recorded (count).

Key question: What type of host do anopheles mosquitoes prefer to feed on, when the host availability is almost 50/50? 

```{r}

rm(list = ls())
# list.files()

```

Load all required packages 

```{r}

library(lme4, warn.conflicts = FALSE) # for generalized linear mixed effect models
library(MASS, warn.conflicts = FALSE) # for generalized models
suppressWarnings(library(tidyverse, warn.conflicts = FALSE)) # for data cleaning and manipulation
library(data.table, v) # import data table package to factor Xticks
library(ggplot2, warn.conflicts = FALSE) # for plotting
library(effects, warn.conflicts = FALSE)
library(ggeffects, warn.conflicts = FALSE)
library(sjPlot, warn.conflicts = FALSE)

```

load data

```{r}


# setwd("C:/Mannu/Projects/Mannu Phd/GLM_metadata")
metadata_df <- read.csv("C:/Mannu/Projects/Mannu Phd/GLM_metadata/_blood_meal_metadata_final.csv")

# factor house hold id column
metadata_df[,'hh_id'] <- as.factor(metadata_df[,'hh_id'])

# drop unused column
metadata_df <- select(metadata_df, -X)

# first print to the console all of the unique values you will need to recode)
print(distinct(metadata_df, host_blood))

# view data
glimpse(metadata_df)

```

Recode all cow written factors as bovine

```{r}


metadata_df$host_blood <- recode(metadata_df$host_blood,
                                      Human_cow = "Human_bovine",
                                      Cow_dog = "Bovine_dog")
metadata_df$host_blood <- factor(metadata_df$host_blood, levels = c("Human", "Bovine", "Dog", 
                                                                    "Human_bovine", "Human_dog", "Bovine_dog"))


print(distinct(metadata_df, host_blood))

```


Exploring the data and visualizing the data

```{r}

# table() is a useful function for understanding the study design
# that produced the data set

print(table(metadata_df$host_blood)) 
print(table(metadata_df$species))

```

```{r}
# # mosquito species blood-fed on vertebrate

table(metadata_df$host_blood, metadata_df$species)

```

```{r}

sumC <- tapply(metadata_df$count, list(metadata_df$host_blood,     metadata_df$species, metadata_df$position, metadata_df$trap_method), sum)
sumC

```


```{r}


meanC <- tapply(metadata_df$count, list(metadata_df$host_blood, metadata_df$species, metadata_df$position), mean)
meanC

```

Explanatory Plotting

```{r}

ggplot(metadata_df, aes(x = host_blood, y = count)) +
  geom_boxplot(aes(fill = host_blood)) + # distinguish mosquito species by color
  geom_dotplot(binaxis="y", binwidth=4, stackdir="center", fill=1, alpha = 0.5) + # plot row data
  stat_summary(fun = mean, colour = "red", geom = "point", # add mean summary to the plot
  shape = 15, size = 4, show.legend = FALSE) + # estimate mean of the raw data
  facet_wrap(~ species) + # divide the data by Host to make two plot for host
  theme_bw() +
  ylab(expression(bold("No. of bloodfed mosquitoes"))) + # name y label
  xlab(expression(bold("Host"))) + # name X label
  theme(text = element_text(size = 18), # adjust font size
        legend.position = "none",
        axis.text.x = element_text(margin = margin(t = 0, r = 0, b = 13, l = 0), angle = 90, vjust = 0.5, hjust = 1)) # create space between

```

```{r}

ggplot(metadata_df, aes(x = host_blood, y = count, fill = host_blood)) + # distinguish host by color
  geom_boxplot(width = 0.5) + 
  geom_dotplot(binaxis="y", binwidth=4, stackdir="center", fill=1, alpha = 0.5) + # plot row data
  stat_summary(fun = mean, colour = "red", geom = "point",
               shape = 15, size = 5,show.legend = FALSE) + # add mean summary
  theme_bw() +
  ylab(expression(bold("No. of bloodfed mosquitoes"))) +
  xlab(expression(bold("Host"))) +
  theme(text = element_text(size = 18), # adjust plot font size
        legend.position = "none",
        axis.text.x = element_text(margin = margin(t = 0, r = 0, b = 13, l = 0), angle = 90, vjust = 0.5, hjust = 1))


```


*Model Fitting*

House hold Id will be treated as the random effects because we aren’t interested in generalizing their effects beyond the experiment.

Host, mosquito species, position, and trap method are treated as fixed/explanatory variables because the purpose of the experiment is to generalize the HOST preference on the number of mosquitoes blood fed on different vertebrates .

The response variable count (no. of mosquitoes blood fed on a particular host), is modeled against the explanatory variables:

(1) Host (different vertebrate which mosquito blooed-fed) - a factor with six levels 

(2) Position (indoor/outdoor collection) - a factor with two levels

(3) Species (An. funestus/An. arabiensis) - a factor with two levels

and INDIVIDUALS (different individuals used in the study) as a random effect

*Null hypothesis*

N0: Anopheles mosquitoes prefer to bite humans than other vertebrate.

Since our response variable is a count, then Poisson GLMM will be a suitable model structure. 

We will <glmer> to fit a number of different models from complicated to simple, and likelihood ration test to test the importance of each variable in the model


```{r}


model1 <- glmer(count ~ host_blood + species + position + (1|hh_id), family = poisson, data = metadata_df)
summary(model1)


# RR <- exp(coef(summary(model1))[,1])
# lower <- exp(coef(summary(model1))[,1] + qnorm(.025)*coef(summary(a1))[,2])
# upper <- exp(coef(summary(model1))[,1] + qnorm(.975)*coef(summary(a1))[,2])
# users_indoors = cbind(RR, lower, upper, pvalue=coef(summary(model1))[,4])


```

function to estimate and test over dispersion

from https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html


```{r}

# Checking for over dispersion 

overdisp_fun <- function(model) {
  rdf <- df.residual(model)
  rp <- residuals(model,type="pearson")
  Pearson.chisq <- sum(rp^2)
  prat <- Pearson.chisq/rdf
  pval <- pchisq(Pearson.chisq, df = rdf, lower.tail = FALSE)
  c(chisq = Pearson.chisq, ratio = prat, rdf = rdf, p = pval)
}
# This can also be done using overdisp.glmer() function in RVAideMemoire package. 
# library(RVAideMemoire) 
# overdisp.glmer(model3)

# but for now, let's use the function we have just created.
# Null hypothesis: There is no overdispersion in model 1

overdisp_fun(model1)

```
There is over dispersion in model1 (ratio 2.5, where 1 is optimal, and P < 0.001 for the null hypothesis no over dispersion)

We can consider refit the model using a negative binomial GLMM (zero inflated model to account for over dispersion)


```{r}


model2 <- glmer.nb(count ~ host_blood + species + position + (1|hh_id), data = metadata_df)
summary(model2)

print(overdisp_fun(model2))

```

Checking whether variables have effect on the model, we will use anova to compare the base model with the new model

Model 3: 
Null hypothesis: Host has no effect, drop host from the model

```{r}

model3.no.host <- glmer.nb(count ~ species + position + (1|hh_id), data = metadata_df)
summary(model3.no.host)

```

```{r}

anova(model2, model3.no.host)


```

From anova, we reject the null hypothesis that host has no effect on the model. we accept the alternative hypothesis that host has effect given LRT: DF = 5, chisq = 50.3, p < 0.001

Model 4: 
Null hypothesis: species has no effect, drop species from the model

```{r}

model4.no.species <- glmer.nb(count ~ host_blood + position + (1|hh_id), data = metadata_df)
summary(model4.no.species)

```

```{r}

anova(model2, model4.no.species)


```

From anova, we reject the null hypothesis that species has no effect on the model. we accept the alternative hypothesis that species has effect given LRT: DF = 1, chisq = 6.9, p < 0.001

Model 5: 
Null hypothesis: position has no effect, drop position from the model

```{r}

model5.no.position <- glmer.nb(count ~ host_blood + species + (1|hh_id), data = metadata_df)
summary(model5.no.position)

```


```{r}

anova(model2, model5.no.position)

```

From anova, we accept the null hypothesis that position has no effect on the model,  given LRT: DF = 1, chisq = 1.9, p > 0.05

Refit the final model without position

```{r}
metadata_df_new = filter(metadata_df, host_blood != "Human_dog" & host_blood != "Bovine_dog")
final.model = glmer.nb(count ~ host_blood + species + (1|hh_id), data = metadata_df_new)
summary(final.model)

```
Inspect model


```{r}

plot(resid(final.model))
hist(resid(final.model))
plot(final.model)

# check the Q-Q plot, to evaluate the assumption of normality of 
# the residuals.
plot_model(final.model, type = "diag") 

# plot(allEffects(final.model)) # plot all model effects
plot(predictorEffects(final.model, ~ host_blood))

```




Exponentiation of the coefficients to get relative ratios, 95 CI values and report results of this final model

```{r}

RR <- exp(coef(summary(final.model))[,1])
lower <- exp(coef(summary(final.model))[,1] + qnorm(.025)*coef(summary(final.model))[,2])
upper <- exp(coef(summary(final.model))[,1] + qnorm(.970)*coef(summary(final.model))[,2])
cbind(RR, lower, upper, pvalue = coef(summary(final.model))[,4])

```

Plotting figures from the model estimates

```{r}

df_hosts <- ggpredict(final.model, terms = c("host_blood"), ci.lvl = 0.95)

ggplot(df_hosts, aes(x = x, y = predicted)) +
  # scale_fill_manual(values = c("gray60")) +
  geom_bar(fill = "lightblue", color = "lightblue", alpha = .4,  stat = "identity", width = 0.6) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.1) +
  theme_bw() +
  ylab(expression(bold("# of bloodfed mosquitoes"))) +
  xlab(expression(bold(" "))) +
  theme(text = element_text(size = 17), # adjust plot font size
        # aspect.ratio = 1/1,
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.text.x = element_text(margin = margin(t = 0, r = 0, b = 13, l = 0), angle = 90, vjust = 0.5, hjust = 1))

df_hosts

```


```{r}

```




